Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-13 ind_files/01_Introduction.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

\section{Idealization and generality in ecology}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations~\citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. Som philosophers of science have recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena. To reiterate once more, the goal of such simple models is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

Vellend has recently argued that conceptual synthesis in community ecology requires ``shifting the emphasis away from an organizational structure based on the useful lines of inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns"~\citep{vellend_theory_2016}. Vellend's assertion is based on the fact that population genetics has managed to come up with reasonably comprehensive theory due to its focus on the abstract `high-level' processes of selection, mutation, drift, and gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources ($R^*$), storage effects, priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table 5.1 in~\citep{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that community ecology `is a mess'. Vellend proposes organizing ecological models according to the `high-level' processes of selection, ecological drift (demographic stochasticity), dispersal, and speciation.

Of course, no such general organization will be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, formulating such (blurry-eyed) general organizing frameworks can be greatly beneficial as an aid to thinking. In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological details but are `general' in the sense of capturing broad features of biological systems (as opposed to the mathematician's notion of `general' in sense of always being true).

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner.

>>>>

\section{Idealization and generality in ecology}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations~\citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. Some philosophers of science have recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena.

To reiterate once more, the goal of such simple models and organizing frameworks is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

A general organization like this will often not be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, formulating such (blurry-eyed) general organizing frameworks can be greatly beneficial as an aid to thinking. This is perhaps best illustrated by the success of statistical mechanics in physics - Statistical mechanics was essentially born from the idea that statistical regularities of systems can emerge without the need for knowing the excruciating details of every single moving part, and indeed, this sort of explicitly `blurry-eyed' thinking was shown to be able to recover the phenomenological laws of thermodynamics from first principles. In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological details but are `general' in the sense of capturing broad features of biological systems (as opposed to the mathematician's notion of `general' in sense of always being true). In a similar spirit, Vellend has recently argued that conceptual synthesis in community ecology requires
``shifting the emphasis away from an organizational structure based on the useful lines of
inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns''~\citep{vellend_theory_2016}. Vellend's assertion is based on the
fact that population genetics has managed to come up  with reasonably comprehensive theory due to its focus on the abstract `high-level’ processes of selection, mutation, drift, and
gene flow (see section below) instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on
specific `low-level' processes such as predation rate, limiting resources (R\textsuperscript{*}), storage effects,
priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table
5.1 in~\cite{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that
community ecology `is a mess'. Vellend instead proposes organizing ecological models according to
 the `high-level’ processes of selection, ecological drift (demographic stochasticity), dispersal,
and speciation, in analogy with population genetics.

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. The mathematical formalism itself is very general and applies equally to the `high level' forces of population genetics as postulated during the Modern Synthesis and the `high level' forces of community ecology as postulated by \cite{vellend_theory_2016}, though I will mostly stick to the population genetics interpretation in my discussions. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner.

### 2023-03-13 ind_files/02_BD_processes.tex
Minor phrasing changes
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

\epigraph{\justifying Somewhere [ ... ] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding}

@@

\section{Description of the process and the Master Equation}
Given a population that can contain up to $m$ different (fixed) kinds of organisms, it can be entirely characterized by specifying the number of organisms of each type (Figure \ref{fig_nD_pop_description}A). Thus, the state of the population at a given time $t$ is an $m$-dimensional \emph{vector} of the form $\mathbf{v} = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of individuals of type $i$.

Given a state $\mathbf{v}(t)$,  we also need to describe how this vector can change over time due to births and deaths (ecology). In this case, a birth or death could result in an individual belonging to one of $m$ different types. Thus, whereas before we had two functions $b(n)$ and $d(n)$ which take in a number as an input, we now require $2m$ functions that take in a vector as an input (Figure \ref{fig_nD_pop_description}B). In other words, for each type $i \in \{1,2,\ldots,m\}$, we must specify a birth rate $b_i(\mathbf{v})$ and a death rate $d_i(\mathbf{v})$. By `rates', we mean that if we know that \emph{either a birth or a death} occurs, then the probability that this event is the birth of an individual of type $i$ is given by
\begin{equation*}
\mathbb{P}[\textrm{ Birth of a type $i$ individual} | \textrm{ something happened }] = \frac{b_i(\mathbf{v})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{v})+d_j(\mathbf{v}))}
\end{equation*}
and the probability that the event is the death of an individual of type $i$ is
\begin{equation*}
\mathbb{P}[\textrm{ Death of a type $i$ individual} | \textrm{ something happened }] = \frac{d_i(\mathbf{v})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{v})+d_j(\mathbf{v}))}
\end{equation*}

\myfig{0.9}{Media/3.2_BD_process_2D.png}{\textbf{Schematic description of a one-dimensional birth-death process}. \textbf{(A)} Consider a population of birds in which individuals are either {\color{red}red} or {\color{blue}blue}. In this case, we have $m=2$, since there are two types of individuals in the population. \textbf{(B)} The state of the system can be described by a vector containing the number of individuals of each discrete type (in this case, the number of red and blue birds in the population). Births and deaths result in changes in the elements of the state vector.}{fig_nD_pop_description}

As before, we can describe the rate of change of $P(\mathbf{v},t)$, the probability of finding the population in a state $\mathbf{v}$ at time $t$, by measuring the inflow and outflow rates. Given a population $\mathbf{v} = [v_1,\ldots,v_{m}]$, the `inflow' is from all populations of the form $[v_1,\ldots,v_{i}-1,\dots,v_{m}]$ through a birth of a type $i$ individual, and from all populations of the form $[v_1,\ldots,v_{i}+1,\dots,v_{m}]$ through the death of a type $i$ individual. Thus, we have the inflow rate
\begin{equation}
\label{nD_rate_in}
\begin{split}
R_{\textrm{in}}(\mathbf{v},t) &= \sum\limits_{j=1}^{m}b_{j}([v_1,\ldots,v_{j}-1,\ldots,v_m])P([v_1,\ldots,v_{j}-1,\ldots,v_m],t) \\
& +\sum\limits_{j=1}^{m}d_{j}([v_1,\ldots,v_{j}+1,\ldots,v_m])P([v_1,\ldots,v_{j}+1,\ldots,v_m],t)
\end{split}
\end{equation}
Outflow is through births and deaths of individuals in the population $\mathbf{v}$ itself, and thus we have:
\begin{equation}
\label{nD_rate_out}
R_{\textrm{out}}(\mathbf{v},t) = \sum\limits_{j=1}^{m}b_{j}(\mathbf{v})P(\mathbf{v},t) + \sum\limits_{j=1}^{m}d_{j}(\mathbf{v})P(\mathbf{v},t)
\end{equation}
As before, we now define step operators, both for notational ease and in anticipation of the system size expansion. Note that now, we need $2m$ step operators. For each $i \in \{1,\ldots,m\}$, let us define two step operators $\mathcal{E}_{i}^{\pm}$ by their action on any function $f([v_1,\ldots,v_m],t)$ as:
\begin{equation}
\label{nD_step_operators}
\mathcal{E}_{i}^{\pm}f([v_1,\ldots,v_m],t) = f([v_1,\ldots,v_i \pm 1, \ldots v_m],t)
\end{equation}
In other words, $\mathcal{E}_{i}^{\pm}$ just changes the population through the addition or removal of one type $i$ individual. We can now the rate of change of $P(\mathbf{v},t)$ as
\begin{equation}
\frac{\partial P}{\partial t}(\mathbf{v},t) = R_{\textrm{in}}(\mathbf{v},t) - R_{\textrm{out}}(\mathbf{v},t)
\end{equation}
Substituting \eqref{nD_rate_in}, \eqref{nD_rate_out}, and \eqref{nD_step_operators}, we obtain:
\begin{equation}
\label{nD_M_eqn}
\frac{\partial P}{\partial t}(\mathbf{v},t) = \sum\limits_{j=1}^{m}\left[(\mathcal{E}_j^{-}-1)b_j(\mathbf{v})P(\mathbf{v},t) + (\mathcal{E}_j^{+}-1)d_j(\mathbf{v})P(\mathbf{v},t)\right]
\end{equation}
This is the master equation of our $m$-dimensional process.

\section{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{v}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{v})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{v})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{B}_t$ is the $m$-dimensional Brownian motion and can be thought of as a vector of independent one-dimensional Brownian motions (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike~\citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation.  Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, we show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]
\end{aligned}
\end{equation}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes and we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. We will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in Chapter \ref{chap_unification}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'.

@@

This is a Price equation for quantities $f_i$ which can vary over time. To obtain the more familiar Price equation seen in textbooks, we can consider time-independent $f_i$, \emph{i.e.} situations in which each $f_i$ is constant over time, and thus changes in $\overline{f}$ are purely due to changes in the composition of the population. For such quantities, we have $\frac{\partial f_i}{\partial t} = 0 \ \forall \ i$ and thus obtain 

@@

, the famous Price equation in continuous time. The first term of the RHS describes the statistical covariance between the quantity $f$ and the fitness $w$. The second term describes `transmission bias' due to mutational effects - The first summation is the `inflow' of $f$ due to mutations, and the second is the `outflow'.

@@

In multi-variable calculus, the directional derivative\footnote{Physicists sometimes use the notation $\partial_{\mathbf{v}}f(\mathbf{x})$ or $\mathbf{v}\cdot\nabla f(\mathbf{x})$ for this object.} $D_{\mathbf{v}}(f(\mathbf{x}))$ of a multidimensional function $f: \mathbb{R}^n \to \mathbb{R}$ along a vector $\mathbf{v}$ is the function defined by:

@@

D_{\mathbf{v}}(f(\mathbf{x})) \coloneqq \sum\limits_{i=1}^{n}\left(\frac{\partial f(\mathbf{x})}{\partial x_i}\right)v_i = \lim_{h \to 0}\frac{f(\mathbf{x}+h\mathbf{v})-f(\mathbf{x})}{h}

>>>>

\epigraph{\justifying Somewhere [...] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding}

@@

\section{Description of the process and the Master Equation}
Given a population that can contain up to $m$ different (fixed) kinds of organisms, it can be entirely characterized by specifying the number of organisms of each type (Figure \ref{fig_nD_pop_description}A). Thus, the state of the population at a given time $t$ is an $m$-dimensional \emph{vector} of the form $\mathbf{n}(t) = [n_1(t),n_2(t),\ldots,n_m(t)]$, where $n_i(t)$ is the number of individuals of type $i$ in the population at time $t$.

Given a state $\mathbf{n}(t)$,  we also need to describe how this vector can change over time due to births and deaths (ecology). In this case, a birth or death could result in an individual belonging to one of $m$ different types. Thus, whereas before we had two functions $b(n)$ and $d(n)$ which take in a number as an input, we now require $2m$ functions that take in a vector as an input (Figure \ref{fig_nD_pop_description}B). In other words, for each type $i \in \{1,2,\ldots,m\}$, we must specify a birth rate $b_i(\mathbf{n})$ and a death rate $d_i(\mathbf{n})$. By `rates', we mean that if we know that \emph{either a birth or a death} occurs, then the probability that this event is the birth of an individual of type $i$ is given by
\begin{equation*}
\mathbb{P}\big[\textrm{ Birth of a type $i$ individual } \big{|} \textrm{ something happened }\big] = \frac{b_i(\mathbf{n})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{n})+d_j(\mathbf{n}))}
\end{equation*}
and the probability that the event is the death of an individual of type $i$ is
\begin{equation*}
\mathbb{P}\big[\textrm{ Death of a type $i$ individual } \big{|} \textrm{ something happened }\big] = \frac{d_i(\mathbf{n})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{n})+d_j(\mathbf{n}))}
\end{equation*}

\myfig{0.9}{Media/3.2_BD_process_2D.png}{\textbf{Schematic description of a multi-dimensional birth-death process}. \textbf{(A)} Consider a population of birds in which individuals are either {\color{red}red} or {\color{blue}blue}. In this case, we have $m=2$, since there are two types of individuals in the population. \textbf{(B)} The state of the system can be described by a vector containing the number of individuals of each discrete type (in this case, the number of red and blue birds in the population). Births and deaths result in changes in the elements of the state vector.}{fig_nD_pop_description}

As before, we can describe the rate of change of $P(\mathbf{n},t)$, the probability of finding the population in a state $\mathbf{n}$ at time $t$, by measuring the inflow and outflow rates. Since the population changes in units of exactly one individual (by definition of a birth-death process; see section \ref{sec_math_background}), we know that these inflow and outflow rates must only involve populations that are a single individual away from our focal population. In other words, for a population $\mathbf{n} = [n_1,\ldots,n_{m}]$, the `inflow' is from all populations of the form $[n_1,\ldots,n_{i}-1,\dots,n_{m}]$ through a birth of a type $i$ individual, and from all populations of the form $[n_1,\ldots,n_{i}+1,\dots,n_{m}]$ through the death of a type $i$ individual. Thus, we have the inflow rate
\begin{equation}
\label{nD_rate_in}
\begin{split}
R_{\textrm{in}}(\mathbf{n},t) &= \sum\limits_{j=1}^{m}b_{j}([n_1,\ldots,n_{j}-1,\ldots,n_m])P([n_1,\ldots,n_{j}-1,\ldots,n_m],t) \\
& +\sum\limits_{j=1}^{m}d_{j}([n_1,\ldots,n_{j}+1,\ldots,n_m])P([n_1,\ldots,n_{j}+1,\ldots,n_m],t)
\end{split}
\end{equation}
Outflow is through births and deaths of individuals in the population $\mathbf{n}$ itself, and thus we have:
\begin{equation}
\label{nD_rate_out}
R_{\textrm{out}}(\mathbf{n},t) = \sum\limits_{j=1}^{m}b_{j}(\mathbf{n})P(\mathbf{n},t) + \sum\limits_{j=1}^{m}d_{j}(\mathbf{n})P(\mathbf{n},t)
\end{equation}
As before, we now define step operators, both for notational ease and in anticipation of the system size expansion. Note that now, we need $2m$ step operators. For each $i \in \{1,\ldots,m\}$, let us define two step operators $\mathcal{E}_{i}^{\pm}$ by their action on any function $f([n_1,\ldots,n_m],t)$ as:
\begin{equation}
\label{nD_step_operators}
\mathcal{E}_{i}^{\pm}f([n_1,\ldots,n_m],t) = f([n_1,\ldots,n_i \pm 1, \ldots n_m],t)
\end{equation}
In other words, $\mathcal{E}_{i}^{\pm}$ just changes the population through the addition or removal of one type $i$ individual. We can now the rate of change of $P(\mathbf{n},t)$ as
\begin{equation}
\frac{\partial P}{\partial t}(\mathbf{n},t) = R_{\textrm{in}}(\mathbf{n},t) - R_{\textrm{out}}(\mathbf{n},t)
\end{equation}
Substituting \eqref{nD_rate_in}, \eqref{nD_rate_out}, and \eqref{nD_step_operators}, we obtain:
\begin{equation}
\label{nD_M_eqn}
\frac{\partial P}{\partial t}(\mathbf{n},t) = \sum\limits_{j=1}^{m}\left[(\mathcal{E}_j^{-}-1)b_j(\mathbf{n})P(\mathbf{n},t) + (\mathcal{E}_j^{+}-1)d_j(\mathbf{n})P(\mathbf{n},t)\right]
\end{equation}
This is the master equation of our $m$-dimensional process.

\section{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{n}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{n})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{n})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{W}_t$ is the $m$-dimensional Wiener process and can be thought of as a vector of independent one-dimensional Wiener processes (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike~\citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation.  Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, we show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes and we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. We will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in Chapter \ref{chap_unification}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'.

@@

This is a Price equation for quantities $f_i$ which can vary over time~\citep{lion_theoretical_2018}. To obtain the more familiar Price equation seen in textbooks, we can consider time-independent $f_i$, \emph{i.e.} situations in which each $f_i$ is constant over time, and thus changes in $\overline{f}$ are purely due to changes in the composition of the population. For such quantities, we have $\frac{\partial f_i}{\partial t} = 0 \ \forall \ i$ and thus obtain the dynamic version of the famous Price equation~\citep{page_unifying_2002,lion_theoretical_2018}:

@@

The first term of the RHS describes the statistical covariance between the quantity $f$ and the fitness $w$. The second term describes `transmission bias' due to mutational effects - The first summation is the `inflow' of $f$ due to mutations, and the second is the `outflow'.

@@

In multi-variable calculus, the directional derivative\footnote{Some authors use the notation $\partial_{\mathbf{v}}f(\mathbf{x})$ or $\mathbf{v}\cdot\nabla f(\mathbf{x})$ for this object.} $D_{\mathbf{v}}(f(\mathbf{x}))$ of a multidimensional function $f: \mathbb{R}^m \to \mathbb{R}$ along a vector $\mathbf{v}$ is the function defined by:

@@

D_{\mathbf{v}}(f(\mathbf{x})) \coloneqq \sum\limits_{i=1}^{m}\left(\frac{\partial f(\mathbf{x})}{\partial x_i}\right)v_i = \lim_{h \to 0}\frac{f(\mathbf{x}+h\mathbf{v})-f(\mathbf{x})}{h}

### 2023-03-13 ind_files/03_field_eqns.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

    \mathbb{P}[\textrm{ Birth with offspring in }A | \textrm{ something happened }] = \frac{1}{\mathcal{N}}\int\limits_{A}b(x|\nu)dx

@@

    \mathbb{P}[\textrm{ Death of an individual in }A | \textrm{ something happened }] = \frac{1}{\mathcal{N}}\int\limits_{A}d(x|\nu)dx

>>>>

    \mathbb{P}\big[\textrm{ Birth with offspring in $A$ } \big{|} \textrm{ something happened }\big] = \frac{1}{\mathcal{N}}\int\limits_{A}b(x|\nu)dx

@@

    \mathbb{P}\big[\textrm{ Death of an individual in $A$ } \big{|} \textrm{ something happened }\big] = \frac{1}{\mathcal{N}}\int\limits_{A}d(x|\nu)dx
